---
title: "Individual Portfolio"
author: "Jaron Dixon"
date: "`r Sys.Date()`"
output:
  github_document: 
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(viridis)  
library(ggridges)
heart_data <- read.csv("heart_attack_dataset_age-1.csv")
```

# ABSTRACT


# BACKGROUND
Every year about 805,000 Americans have a heart attack. (Tsao et al., 2023) A heart attack can pose pain from a mild discomfort to severe crushing pain. (Kulkarni, 2025) This health risk has become like a ticking bomb being able to strike anyone. With this disease that has the ability to strike a variety of different individuals, research has begun to try and predict risk factors or conditions that can cause an individual to be more susceptible to having a heart attack. Long-term or chronic stress can lead to higher levels of inflammation in the body which leads to more plague buildup in the arteries. (Katella, 2024) Additionally, stress also causes hormones such as adrenaline. (Katella, 2024) Adrenaline increases mental alertness and the heart beats faster and raises blood pressure. (Katella, 2024) Prolonged stress leads to heart damage. 

With this in mind we turn to what might cause an individual to have prolonged stress. Specifically, undergraduate students often experience stress related to adapting to higher education and managing coursework. (Pérez-Jorge, 2025) Postgraduate students also exhibit stress about advanced research, thesis completion, and balancing academic work and professional responsibilities. (Pérez-Jorge, 2025) When it comes to pursuing higher education there are a lot of stressful responsibilities for individuals. This could be a cause of having a risk of heart attacks because of the impact of the stress on the body. 

Another factor that might contribute to a higher risk of heart attacks is gender. Researchers have found that men are twice as likely to have a heart attack than women. (Harvard Health Publishing, 2016) There is also some suspicion that hormones in women before menopause cause more defense for women against heart attacks. (Harvard Health Publishing, 2016) 

Knowing this information brings a lot of question of how much factors like stress, gender, or education could lead to an individual being more likely to have a heart attack. 


# STUDY QUESTION and HYPOTHESIS


## Questions  

Do certain factors cause a increased chance in having a heart attack?

## Hypothesis 

If the person is male, have high stress level, and have a post-graduate education level, then we expect to see an increase in amounts of heart attacks.
  
## Prediction
We expect to see people who are male, that have a post graduate degree, and who are under high stress to experience the highest chance of obtaining a heart attack.

# Methods

# Analyses

## Figures

```{r Heart Attack Vs Education Level lollipop Plot, eval=TRUE}

# Summarize counts
edu_counts <- heart_data %>%
  group_by(EducationLevel, Outcome) %>%
  summarise(count = n(), .groups = "drop")

# lollipop chart
ggplot(edu_counts, aes(x = EducationLevel, y = count, color = Outcome)) +
  geom_segment(aes(x = EducationLevel, xend = EducationLevel, y = 0, yend = count), linewidth = 1.2) +
  geom_point(size = 6) +
  scale_color_manual(values = c("No Heart Attack" = "skyblue", "Heart Attack" = "tomato")) +
  labs(
    title = "Heart Attack Outcomes by Education Level",
    x = "Education Level",
    y = "Count of Individuals",
    color = "Heart Attack Outcome"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

```{r Heart Attack vs StressLevel Violin Plot, eval=TRUE}
ggplot(heart_data, aes(x = Outcome, y = StressLevel, fill = Outcome)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white", alpha = 0.5) +
  labs(
    title = "Stress Levels by Heart Attack Outcome",
    x = "Heart Attack Outcome",
    y = "Stress Level (1–9)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "none"
  )

```
```{r Stress vs Heart Attack Ridge Plot, eval=TRUE }

ggplot(heart_data, aes(x = StressLevel, y = Outcome, fill = Outcome)) +
  geom_density_ridges(alpha = 0.7, color = "black", scale = 1) +
  scale_fill_manual(values = c("No Heart Attack" = "skyblue", "Heart Attack" = "tomato")) +
  labs(
    title = "Distribution of Stress Levels by Heart Attack Outcome",
    x = "Stress Level (1–9)",
    y = "Heart Attack Outcome",
    fill = "Outcome"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "none",
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  )
```


```{r Gender vs Heart Attack Heat Plot, eval=TRUE}

# Summarize counts
gender_counts <- heart_data %>%
  group_by(Gender, Outcome) %>%
  summarise(count = n(), .groups = "drop")

#Heat map
ggplot(gender_counts, aes(x = Gender, y = Outcome, fill = count)) +
  geom_tile(color = "white", size = 0.8) +  # white borders between tiles
  geom_text(aes(label = count), color = "white", size = 6, fontface = "bold") +
  scale_fill_viridis(option = "C", direction = -1) +  # smooth and modern gradient
  labs(
    title = "Heart Attack Outcomes by Gender",
    x = "Gender",
    y = "Heart Attack Outcome",
    fill = "Count"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 20),
    axis.text = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 16),
    legend.title = element_text(face = "bold", size = 14),
    legend.text = element_text(size = 12)
  )
```

```{r Stress Levels, Education Level, Gender and Heart Attack combined plot, eval=TRUE}

ggplot(heart_data, aes(
  x = EducationLevel,
  y = StressLevel,
  color = Outcome,       
  alpha = Outcome,  
  shape = Gender
)) +
  geom_jitter(width = 0.3, size = 4) +
  scale_alpha_manual(values = c("No Heart Attack" = .9, "Heart Attack" = 1)) +
  scale_shape_manual(values = c("Female" = 1, "Male" = 16)) +  # hollow vs filled
  scale_color_manual(values = c("No Heart Attack" = "lightblue3", "Heart Attack" = "indianred2")) + 
  labs(
    title = "Stress Levels by Education, Gender, and Heart Attack Outcome",
    x = "Education Level",
    y = "Stress Level (1–9)",
    color = "Heart Attack Outcome",
    alpha = "Heart Attack Outcome",
    shape = "Gender"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

```

## Statistical Analysis

```{r Stress Levels vs Heart Attack Logistic Regression}
# Load necessary libraries
library(ggplot2)
library(broom)  # for tidy model output

# Ensure Outcome is a factor with the correct reference level
heart_data$Outcome <- factor(heart_data$Outcome, levels = c("No Heart Attack", "Heart Attack"))

# Fit logistic regression model
stress_model <- glm(Outcome ~ StressLevel, data = heart_data, family = binomial)

# View model summary
summary(stress_model)

# Calculate Odds Ratios and 95% Confidence Intervals
odds_ratios <- exp(cbind(OR = coef(stress_model), confint(stress_model)))
print(odds_ratios)

# Extract tidy model output for p-value
model_stats <- tidy(stress_model)
p_value <- model_stats$p.value[2]  # p-value for StressLevel
cat("P-value for Stress Level:", round(p_value, 4), "\n")

# Create data for predicted probabilities
pred_data <- data.frame(
  StressLevel = seq(min(heart_data$StressLevel),
                    max(heart_data$StressLevel), length.out = 100)
)
pred_data$PredictedProb <- predict(stress_model, newdata = pred_data, type = "response")

# Plot predicted probability curve with dotted points
ggplot() +
  # Add raw data points (dotted)
  geom_jitter(data = heart_data, aes(x = StressLevel, y = as.numeric(Outcome) - 1),
              width = 0.15, alpha = 0.5, shape = 21, color = "black", fill = "grey80") +
  # Add predicted probability curve
  geom_line(data = pred_data, aes(x = StressLevel, y = PredictedProb),
            color = "red", size = 1.2) +
  # Rug for distribution
  geom_rug(data = heart_data, aes(x = StressLevel), sides = "b", alpha = 0.3) +
  # Labels
  labs(
    title = paste0(
      "Predicted Probability of Heart Attack by Stress Level\n",
      "p-value = ", round(p_value, 4)
    ),
    x = "Stress Level (1–9)",
    y = "Predicted Probability of Heart Attack"
  ) +
  theme_minimal(base_size = 14)
```



```{r Education Level vs Heart Attack Chi Squared}
# Create a contingency table of counts
edu_table <- table(heart_data$EducationLevel, heart_data$Outcome)
print("Contingency Table:")
print(edu_table)

# Perform Chi-Square Test
chi_result <- chisq.test(edu_table)
print("Chi-Square Test Results:")
print(chi_result)
```

```{r Gender vs Heart Attack Chi Squared}
# Create contingency table
gender_table <- table(heart_data$Gender, heart_data$Outcome)
print("Contingency Table:")
print(gender_table)

# Run Chi-Square Test
chi_gender <- chisq.test(gender_table)
print("Chi-Square Test Results:")
print(chi_gender)

# Display expected counts
print("Expected Counts:")
print(chi_gender$expected)

# If Chi-Square assumptions are violated (expected counts < 5), run Fisher's Exact Test
if(any(chi_gender$expected < 5)) {
  print("Expected counts too low — running Fisher's Exact Test instead:")
  fisher_result <- fisher.test(gender_table)
  print(fisher_result)
}
```

```{r Stress Level, Gender, Education Level vs Heart Attack Statistics}

```


# DISCUSSION


## Interpretation 1


# CONCLUSION


# REFERENCES     

1. Harvard Health Publishing. “Throughout Life, Heart Attacks Are Twice as Common in Men than Women - Harvard Health.” Harvard Health, Harvard Health, 8 Nov. 2016, www.health.harvard.edu/heart-health/throughout-life-heart-attacks-are-twice-as-common-in-men-than-women.
2.Katella, Kathy. “Yes, Stress Can Hurt Your Heart: 3 Things to Know.” Yale Medicine, 12 Feb. 2024, www.yalemedicine.org/news/stress-affects-your-heart.
3. Kulkarni, Anandita. “What Does a Heart Attack Feel Like? 8 Warning Signs You Shouldn’t Ignore.” @Bswhealth, 2025, www.bswhealth.com/blog/what-does-a-heart-attack-feel-like-8-warning-signs.
4. Panday, Ankush. “Heart Attack Prediction in United States.” Kaggle.com, 2025, www.kaggle.com/datasets/ankushpanday2/heart-attack-prediction-in-united-states/data. Accessed 6 Nov. 2025.
5. Pérez-Jorge, David, et al. “Examining the Effects of Academic Stress on Student Well-Being in Higher Education.” Humanities and Social Sciences Communications, vol. 12, no. 1, 28 Mar. 2025, pp. 1–13, www.nature.com/articles/s41599-025-04698-y, https://doi.org/10.1057/s41599-025-04698-y.
6. Tsao, Connie W., et al. “Heart Disease and Stroke Statistics—2023 Update: A Report from the American Heart Association.” Circulation, vol. 147, no. 8, 25 Jan. 2023, www.ahajournals.org/doi/10.1161/CIR.0000000000001123, https://doi.org/10.1161/cir.0000000000001123.




