---
title: "Individual Portfolio"
author: "Jaron Dixon"
date: "`r Sys.Date()`"
output:
  github_document: 
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# ABSTRACT


# BACKGROUND


# STUDY QUESTION and HYPOTHESIS


## Questions  

Does the education level of people affect the frequency/likelihood of getting a heart attack due to stress? 

## Hypothesis 

If the person has a higher education level, then they are likely to be more prone to having a heart attack due to higher amounts of stress.
  
## Prediction


# Possible Visualization Method

What we have found out for our statistical tests is a few things. Since we are comparing multiple groups, here are the types of tests that we can do for each one:

Education level and Stress: ANOVA test

Education Level and Heart Attack:  Chi Squared Test of Independence

Stress Levels and Heart Attack: T-Test

Education Level, Stress Levels, and Heart Attack: Binary Logistic Regression

## Assignment 10 Figure

```{r eval=FALSE}
library(ggplot2)

heart_data <- read.csv("heart_attack_dataset_new.csv")

ggplot(heart_data, aes(x = EducationLevel, fill = Outcome)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Heart Attack Outcomes by Education Level",
    x = "Education Level",
    y = "Count of Individuals",
    fill = "Outcome"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "right"
  )

```



# METHODS


## Analysis 1


# DISCUSSION


## Interpretation 1


# CONCLUSION


# REFERENCES     

1. Ankit. (2025). Heart Attack Prediction in United States [Data set]. Kaggle. https://doi.org/10.34740/KAGGLE/DSV/10806451



